A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
Minor.

If $t$ is a selector term %of the form
$\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$
for each $i$ and thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\interp{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
$\smash{\const s^k_{\negvthinspace j}( u )}$ and $\smash{\const s^k_{\negvthinspace j}( u' )}$
are in distinct equivalence classes only if $u$ and $u'$ are in distinct equivalence classes.
Due to the induction hypothesis and Lemma~\ref{lem:interpretation-completion}\PART{3}, $\M( u ) \neq \M( u' )$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\M(\const s^k_{\negvthinspace j})( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes:
Since quantified formulas may entail cardinality constraints on uninterpreted types $\ty{u}$,
we may no longer assume that they are of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus presented in this paper is \relax{incomplete},
due to the overly restrictive preconditions on \rn{Split} and \rn{Single}.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete when quantified formulas are present.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, consider the $\ty{stream}_{\,\ty{u}}$ codatatype, where $\ty{u}$ is uninterpreted.
This type has cardinality precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y \mathrel\vee s \teq t$ to $F$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al. \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance because the alternative introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
There are not yet many benchmarks available, since codatatypes are a
recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. We
included all the theory files from the Isabelle distribution (Distro, 879 goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974 goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the support
for datatypes, the benchmarks were complemented by theories about lists
and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://www21.in.tum.de/~blanchet/codp_data.tgz}}

>>>>

If $t$ is a selector term %of the form
$\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\interp{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$
for each $i$ and thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\interp{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\interp{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
any term of the form $\smash{\const s^k_{\negvthinspace j}( u' )}$ not occurring in $\ec{t}$ 
is such that $u$ and $u'$ are in distinct equivalence classes.
Due to the induction hypothesis and Lemma~\ref{lem:interpretation-completion}\PART{3}, $\M( u ) \neq \M( u' )$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\M(\const s^k_{\negvthinspace j})( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes:
Since quantified formulas may entail cardinality constraints on uninterpreted types $\ty{u}$,
we may no longer assume that they are of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus presented in this paper is \relax{incomplete}
since it may fail to recognize cases when the preconditions of \rn{Split} and \rn{Single} hold.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete when quantified formulas are present.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, consider the $\ty{stream}_{\,\ty{u}}$ codatatype, where $\ty{u}$ is uninterpreted.
This type has cardinality precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al. \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance since it introduces terms less eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
There are not yet many benchmarks available, since codatatypes are a
recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.} We
included all the theory files from the Isabelle distribution (Distro, 879 goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974 goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the support
for datatypes, the benchmarks were complemented by theories about lists
and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://www21.in.tum.de/~blanchet/codp_data.tgz}}
